---
title: "rvarsim: Variant Simulation with HGVS Notation"
author: "Liu Sun"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{rvarsim: Variant Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

### Motivation

Variant interpretation workflows frequently require converting,
validating, and simulating genetic variants across different
representation formats. While Bioconductor provides excellent
infrastructure for working with genomic ranges (*GenomicRanges*,
*GenomicFeatures*) and variant annotation (*VariantAnnotation*,
*ensemblVEP*), no existing Bioconductor package offers a unified
toolkit for **HGVS (Human Genome Variation Society)** nomenclature
operations.

`rvarsim` fills this gap by providing:

- **Variant simulation**: exhaustive generation of all possible
  single nucleotide variants across **MANE** (Matched Annotation
  from NCBI and EMBL-EBI) Select transcripts — the jointly curated,
  one-transcript-per-gene reference set;
- **HGVS parsing and validation**: decomposing HGVS strings (e.g.,
  `NM_000546.6:c.215C>G`) into structured R objects;
- **Format conversion**: bidirectional mapping between HGVS, VCF,
  and **SPDI** (Sequence Position Deletion Insertion, NCBI's 0-based
  interbase format);
- **Transcription mapping**: converting between genomic (`g.`) and
  coding (`c.`) coordinates via transcript annotation;
- **Translation and backtranslation**: inferring protein consequences
  from coding variants, and enumerating possible coding variants
  from observed amino acid changes.

### Comparison with existing tools

| Feature | VariantAnnotation | ensemblVEP | rvarsim |
|---|---|---|---|
| HGVS parsing into R objects | Limited | No | Yes |
| HGVS validation | No | No | Yes |
| Exhaustive SNV simulation | No | No | Yes |
| HGVS ↔ VCF ↔ SPDI conversion | VCF only | VCF only | Yes |
| c. ↔ g. coordinate mapping | Partial | No | Yes |
| Translation to p. notation | No | Limited | Yes |
| Backtranslation (p. → c.) | No | No | Yes |
| Variant extraction from alignment | No | No | Yes |
| Assembly liftover | Via *rtracklayer* | No | Yes |

### Terminology

- **MANE Select**: Matched Annotation from NCBI and EMBL-EBI. A
  jointly curated set of one representative transcript per
  protein-coding gene. MANE Select transcripts are identified by
  RefSeq `NM_` accessions and serve as the standard reference for
  clinical variant reporting.
- **HGVS**: Human Genome Variation Society nomenclature. A
  standardized system for describing sequence variants using
  reference sequences and coordinate systems. Coding DNA notation
  uses `c.` prefix (e.g., `c.215C>G`), genomic uses `g.`, and
  protein uses `p.`.
- **SPDI**: Sequence Position Deletion Insertion. NCBI's variation
  representation using 0-based interbase coordinates (e.g.,
  `NC_000001.11:123455:A:G`). More computationally tractable than
  HGVS for   database storage and comparison.

## Installation

`rvarsim` is a Bioconductor package. Install it with
*BiocManager*:

```{r install, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("rvarsim")
```

Load the package:

```{r load-pkg}
library(rvarsim)
```

## Parsing and Validating HGVS Variants

HGVS strings encode variant positions, reference alleles, and
alternate alleles in a compact but complex grammar. `parse_hgvs()`
decomposes an HGVS string into a structured R list with named
components for programmatic access. This is useful when reading
variants from clinical reports, databases like ClinVar, or
user-provided input.

The code chunk below parses several HGVS variants representing
different regions (CDS, UTR, intronic), validates them, and
inspects their parsed structure.

```{r parse}
# Substitution in coding sequence
cds <- parse_hgvs("NM_000546.6:c.215C>G")[[1]]
cds$type          # "substitution"
cds$reference     # "C"
cds$alternate     # "G"
cds$position$start # 215

# 5' UTR and 3' UTR positions
five_utr <- parse_hgvs("NM_000314.8:c.-14G>A")[[1]]
three_utr <- parse_hgvs("NM_001126114.2:c.*32T>C")[[1]]
five_utr$position$is_five_utr   # TRUE
three_utr$position$is_three_utr # TRUE

# Intronic splice donor and acceptor
donor <- parse_hgvs("NM_000546.6:c.453+1G>T")[[1]]
acceptor <- parse_hgvs("NM_000546.6:c.612-2A>G")[[1]]
donor$position$offset_start     # 1
acceptor$position$offset_start  # -2

# Genomic notation
genomic <- parse_hgvs("NC_000001.11:g.123456A>G")[[1]]
genomic$notation  # "g"
```

`is_valid_hgvs()` performs syntactic and semantic validation of
HGVS strings, returning `TRUE` when a variant conforms to HGVS
grammar and is biologically plausible:

```{r validate}
is_valid_hgvs("NM_000546.6:c.215C>G")  # valid
is_valid_hgvs("NM_000546.6:c.99999A>G") # invalid position
is_valid_hgvs("garbage string")        # unparseable
```

## Format Conversion: HGVS ↔ VCF ↔ SPDI

Variant data moves between systems in different formats. Genomic
analyses typically use **VCF** (Variant Call Format), NCBI
databases use **SPDI**, and clinical reports use **HGVS**. `rvarsim`
provides bidirectional conversion between these formats.

The following chunk demonstrates HGVS-to-VCF and HGVS-to-SPDI
conversion, then verifies round-trip fidelity:

```{r convert}
# HGVS to VCF
vcf <- hgvs_to_vcf("NC_000001.11:g.123456A>G")
print(vcf)

# VCF back to HGVS
vcf_row <- data.frame(CHROM = "1", POS = 123456L,
                      REF = "A", ALT = "G")
vcf_to_hgvs(vcf_row, "GRCh38")

# SPDI round-trip
spdi <- hgvs_to_spdi("NC_000001.11:g.123456A>G")
cat("SPDI:", spdi, "\n")
spdi_to_hgvs(spdi)
```

## Variant Extraction from Sequences

When comparing a reference sequence against an observed
(variant) sequence, `extract_hgvs()` automatically identifies
substitutions, insertions, and deletions and describes them in
HGVS notation. It uses pairwise alignment (Needleman-Wunsch
by default) and returns one HGVS string per detected difference.

This example uses self-contained nucleotide strings to
demonstrate the function without external data:

```{r extract}
# Substitution: C → T at position 4
extract_hgvs("ATGCGT", "ATGCTT", "NM_000546.6", "c", 1)

# Insertion: extra AGA between positions 3 and 4
extract_hgvs("ATGTAG", "ATGAGATAG", "NM_000546.6", "c", 1)

# Deletion: missing G at position 5
extract_hgvs("ATGCGT", "ATGCT", "NM_000546.6", "c", 1)
```

## Normalization

HGVS normalization applies standard rules to produce a canonical
representation: trimming common prefixes/suffixes from deletion-
insertion variants, rewriting insertions as duplications where
applicable, and 3' shifting indels in repetitive regions.

```{r normalize}
# Identity (substitutions are already canonical)
normalize_hgvs("NM_000546.6:c.215C>G")

# Common affix trimming (shared prefix/suffix removed from delins)
normalize_hgvs("NM_000546.6:c.215_217delinsTAG")
```

## Simulating Variants (requires genome data)

The core innovation of `rvarsim` is exhaustive SNV simulation:
for every position across CDS, UTR, and splice site regions of
MANE Select transcripts, all three possible alternative alleles
are generated and formatted in HGVS notation.

This requires both an EnsDb (transcript annotation) and a BSgenome
(reference sequence) **with matching seqlevels styles**. For
example, `EnsDb.Hsapiens.v86` uses UCSC-style chromosome names
(`chr1`) while `BSgenome.Hsapiens.UCSC.hg38` also uses UCSC style,
ensuring compatibility.

```{r simulate, eval=FALSE}
library(EnsDb.Hsapiens.v86)
library(BSgenome.Hsapiens.UCSC.hg38)

# Step 1: Fetch MANE Select transcripts
mane <- fetch_mane_txdb(EnsDb.Hsapiens.v86)
# Step 2: Extract CDS, UTR, splice site structure
struct <- get_transcript_structure(mane, "ENST00000357654")
# Step 3: Generate all possible SNVs
vars <- generate_variants(struct, BSgenome.Hsapiens.UCSC.hg38,
                           regions = c("cds", "splice_site"))
# Step 4: Format in HGVS notation
result <- format_hgvs(vars)
head(result[, c("region", "genomic_ref",
                "genomic_alt", "hgvs_c")])
```

The all-in-one `simulate_variants()` wrapper combines these steps
with progress reporting:

```{r wrapper, eval=FALSE}
result <- simulate_variants(
    txdb     = EnsDb.Hsapiens.v86,
    bsgenome = BSgenome.Hsapiens.UCSC.hg38,
    transcript_ids = "ENST00000357654",
    regions  = c("cds", "splice_site")
)
```

## Genomic-to-Coding Coordinate Mapping

Variants are often reported in genomic (`g.`) coordinates but
interpreted in transcript (`c.`) coordinates. `c_to_g()` and
`g_to_c()` map between these systems using the exon/CDS structure
from a TxDb. These functions handle intronic offsets (`c.453+1`),
UTR positions (`c.-14`, `c.*32`), and strand orientation
automatically.

```{r transcribe, eval=FALSE}
txdb <- EnsDb.Hsapiens.v86
bsg  <- BSgenome.Hsapiens.UCSC.hg38

# Coding → Genomic
c_to_g("ENST00000357654:c.215C>G", txdb, bsg)

# Genomic → Coding
g_to_c("1:g.7577120C>G", txdb, bsg, "ENST00000357654")
```

## Translation and Backtranslation

`translate_hgvs()` infers protein (`p.`) consequences from coding
variants using the standard genetic code. It detects missense,
nonsense, silent, frameshift, and inframe indel consequences.

`backtranslate_hgvs()` performs the reverse: given an observed
protein change (e.g., `p.Arg215Gly`), it enumerates all possible
coding variants by walking the genetic code's codon mappings.
This is useful for interpreting proteomics data.

```{r translate, eval=FALSE}
txdb <- EnsDb.Hsapiens.v86
bsg  <- BSgenome.Hsapiens.UCSC.hg38

# Translate a coding variant to protein consequence
translate_hgvs("ENST00000357654:c.215C>G", txdb, bsg)

# Enumerate coding variants that produce p.Arg215Gly
backtranslate_hgvs("p.Arg215Gly", txdb, bsg)
```

## Liftover Between Assemblies

`liftover_hgvs()` maps genomic variants between assemblies
(e.g., hg19 → hg38) using UCSC chain files. `liftover_c_hgvs()`
performs full c. → g. → liftover → c. conversion in a single
call, preserving coding notation across assemblies.

```{r liftover, eval=FALSE}
library(BSgenome.Hsapiens.UCSC.hg38)

# Liftover a genomic variant from hg19 to hg38
liftover_hgvs("NC_000001.10:g.123456A>G",
              "hg19ToHg38.over.chain",
              BSgenome.Hsapiens.UCSC.hg38)
```

## Interoperability with Other Bioconductor Packages

The returned data structures are designed to integrate with
standard Bioconductor workflows. For example, a variant table
from `simulate_variants()` can be converted to a `GRanges` for
use with *VariantAnnotation* or overlap analysis:

```{r interoperate}
# Simulated variant table → GRanges
vars <- data.frame(
    transcript_id = "NM_TEST",
    genomic_pos   = c(215L, 453L),
    genomic_ref   = c("C", "G"),
    genomic_alt   = c("G", "T"),
    stringsAsFactors = FALSE
)
formatted <- format_hgvs(vars)
print(formatted[, c("genomic_pos", "genomic_ref",
                     "genomic_alt", "hgvs_c")])
```

## References

1. Morales J, Pujar S, Loveland JE, et al. A joint NCBI and
   EMBL-EBI transcript set for clinical genomics and research.
   *Nature*, 2022.
2. den Dunnen JT, Dalgleish R, Maglott DR, et al. HGVS
   recommendations for the description of sequence variants:
   2016 update. *Human Mutation*, 2016; 37(6):564-569.
3. Holmes JB, Moyer E, Phan L, et al. SPDI: data model for
   variants and applications at NCBI. *Bioinformatics*, 2020;
   36(6):1902-1907.

## Session Information

```{r sessionInfo}
sessionInfo()
```
